---
title: "Suspicious Stops and Arrests, Adjusted for Neighborhood Demographics"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Load data

```{r load}
library(dplyr)
library(ggplot2)
source("R/clean_stop_data.R")
source("R/classify_stops.R")

data <- clean_stop_data()
suspicious <- flag_suspicious_stops(data) %>% classify_outcome()
```

## Tie each stop to a census tract

`STOP_LOCATION_BLOCK` is a block-level address (e.g. `"6600 BLOCK OF 3RD
STREET NW..."`), not a lat/lon. The unique blocks behind these suspicious
stops were geocoded once via the US Census Bureau's free batch geocoder
(see `R/geocode_utils.R`) and cached to `data/geocoded_blocks.csv`, which
returns a census tract GEOID directly for each matched address. Not every
block-level address geocodes cleanly (some are missing a city/zip suffix,
some are intersections the geocoder can't resolve), so this join drops a
minority of stops.

```{r join-tract}
geocoded <- read.csv("data/geocoded_blocks.csv", colClasses = c(GEOID = "character")) %>%
  select(STOP_LOCATION_BLOCK, GEOID)

suspicious <- suspicious %>% left_join(geocoded, by = "STOP_LOCATION_BLOCK")

match_rate <- mean(!is.na(suspicious$GEOID))
cat(sprintf(
  "%d / %d suspicious stops (%.1f%%) matched to a census tract\n",
  sum(!is.na(suspicious$GEOID)), nrow(suspicious), 100 * match_rate
))
```

## Load tract-level population by race

The DC census tract ACS extract in `data/census_data/` reliably populates
only two race columns — White (`DP05_0037E`) and Black (`DP05_0038E`); the
Asian, "two or more races", and Hispanic/Latino columns are almost entirely
zero (a gap in the source file itself), so this analysis is limited to a
Black/White comparison, matching the convention already used in
`explore/explore_acs_dc_data.R`.

```{r load-tracts}
tracts <- read.csv("data/census_data/ACS_5-Year_Demographic_Characteristics_DC_Census_Tract.csv",
  colClasses = c(GEOID = "character")
) %>%
  transmute(
    GEOID,
    total_pop = DP05_0001E,
    white_pop = DP05_0037E,
    black_pop = DP05_0038E,
    black_pct = black_pop / total_pop
  )
```

## Citywide stop and arrest rates, adjusted for population by race

This compares the race of the *person stopped* against DC's citywide Black
and White population — it answers "relative to how many Black/White
residents DC has, how often are Black/White people stopped or arrested in
these suspicious stops?" It does not depend on the tract geocoding (every
suspicious stop already has an `ETHNICITY` field), so it's unaffected by
the geocoding match-rate limitation above.

```{r citywide-rates}
citywide_pop <- tracts %>% summarise(white_pop = sum(white_pop), black_pop = sum(black_pop))

citywide_rates <- suspicious %>%
  filter(ETHNICITY %in% c("Black", "White")) %>%
  group_by(ETHNICITY) %>%
  summarise(stops = n(), arrests = sum(OUTCOME == "Arrest"), .groups = "drop") %>%
  mutate(
    population = ifelse(ETHNICITY == "Black", citywide_pop$black_pop, citywide_pop$white_pop),
    `Stop rate per 1,000 residents` = stops / population * 1000,
    `Arrest rate per 1,000 residents` = arrests / population * 1000
  )

citywide_rates
```

```{r plot-citywide-rates}
citywide_colors <- c("Black" = "#2a78d6", "White" = "#eb6834")

citywide_long <- citywide_rates %>%
  tidyr::pivot_longer(
    cols = c(`Stop rate per 1,000 residents`, `Arrest rate per 1,000 residents`),
    names_to = "metric", values_to = "rate"
  ) %>%
  mutate(metric = factor(metric, levels = c("Stop rate per 1,000 residents", "Arrest rate per 1,000 residents")))

citywide_plot <- citywide_long %>%
  ggplot(aes(x = metric, y = rate, fill = ETHNICITY)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = round(rate, 1)), position = position_dodge(width = 0.6), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = citywide_colors, name = "Ethnicity") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    title = "Suspicious Stop and Arrest Rates, Black vs. White Residents",
    subtitle = "Per 1,000 DC residents of that race (ACS population denominator)",
    x = NULL,
    y = "Rate per 1,000 residents"
  )

citywide_plot
ggsave("plots/demographics_citywide_rates.png", plot = citywide_plot, width = 8, height = 5.5)
```

## Tract-level stop intensity vs. neighborhood racial composition

This is a different, complementary question: regardless of who is stopped,
*where* do suspicious stops concentrate, and does that track a
neighborhood's racial makeup? Each point is a census tract: the number of
suspicious stops located in that tract (any ethnicity), divided by the
tract's total population. Tracts with fewer than 100 residents are dropped
(e.g. GEOID 11001980000 is a mostly-water, non-residential tract with only
17 residents, whose tiny denominator turns a handful of stops into a
meaningless 400+ per-1,000 "rate").

```{r tract-summary}
tract_summary <- suspicious %>%
  filter(!is.na(GEOID)) %>%
  group_by(GEOID) %>%
  summarise(n_stops = n(), n_arrests = sum(OUTCOME == "Arrest"), .groups = "drop") %>%
  inner_join(tracts, by = "GEOID") %>%
  filter(total_pop >= 100) %>%
  mutate(
    stops_per_1k = n_stops / total_pop * 1000,
    arrests_per_1k = n_arrests / total_pop * 1000
  )

cat(sprintf("%d of %d DC census tracts had at least one geocoded suspicious stop\n", nrow(tract_summary), nrow(tracts)))
cat(sprintf(
  "Correlation between tract %% Black population and stops per capita: %.2f\n",
  cor(tract_summary$black_pct, tract_summary$stops_per_1k)
))
```

```{r plot-tract-scatter}
tract_scatter <- tract_summary %>%
  ggplot(aes(x = black_pct, y = stops_per_1k)) +
  geom_point(color = "#2a78d6", size = 2, alpha = 0.7) +
  geom_smooth(method = "lm", color = "#eb6834", se = TRUE, linewidth = 0.9) +
  scale_x_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(plot.subtitle = element_text(size = 9)) +
  labs(
    title = "Suspicious Stops per Capita vs. Tract Black Population Share",
    subtitle = stringr::str_wrap(
      sprintf(
        "Each point is one DC census tract; moderate positive correlation (r = %.2f, R-squared = %.2f)",
        cor(tract_summary$black_pct, tract_summary$stops_per_1k),
        summary(lm(stops_per_1k ~ black_pct, data = tract_summary))$r.squared
      ),
      width = 90
    ),
    x = "Tract population that is Black",
    y = "Suspicious stops per 1,000 residents"
  )

tract_scatter
ggsave("plots/demographics_tract_scatter.png", plot = tract_scatter, width = 8, height = 5.5)
```

Neighborhoods with a higher Black population share do see moderately more
suspicious stops per capita (r of about 0.42), but that relationship alone
only explains about 17% of the tract-to-tract variation (R-squared of
about 0.17) -- there's substantial scatter, and plenty of majority-Black tracts sit at very low
per-capita stop rates while a few low-Black-population tracts run high.
That partial geographic relationship is far too small on its own to
account for the citywide gap above, where Black residents are stopped at
roughly `r round(citywide_rates$"Stop rate per 1,000 residents"[citywide_rates$ETHNICITY == "Black"] / citywide_rates$"Stop rate per 1,000 residents"[citywide_rates$ETHNICITY == "White"])`x
the per-capita rate of White residents and arrested at roughly
`r round(citywide_rates$"Arrest rate per 1,000 residents"[citywide_rates$ETHNICITY == "Black"] / citywide_rates$"Arrest rate per 1,000 residents"[citywide_rates$ETHNICITY == "White"])`x
the rate. Put together, this suggests the disparity isn't
mainly a story of "stops happen to be concentrated in heavily-Black
neighborhoods" — it's that Black individuals are stopped and arrested at
far higher rates relative to their population however you slice it,
including within tracts that aren't especially Black.

## Map: arrests per capita by tract

```{r plot-tract-map}
if (requireNamespace("sf", quietly = TRUE)) {
  library(sf)
  source("R/geo_viz_utils.R")

  dc_tracts_geo <- read_census_tract_geo() %>%
    mutate(GEOID = as.character(GEOID)) %>%
    left_join(tract_summary %>% select(GEOID, arrests_per_1k), by = "GEOID")

  tract_map <- ggplot(dc_tracts_geo) +
    geom_sf(aes(fill = arrests_per_1k), color = "white", linewidth = 0.1) +
    scale_fill_gradient(low = "#cde2fb", high = "#0d366b", na.value = "grey90", name = "Arrests\nper 1,000") +
    theme_minimal() +
    labs(title = "Suspicious-Stop Arrests per 1,000 Residents, by Census Tract")

  tract_map
  ggsave("plots/demographics_tract_map.png", plot = tract_map, width = 8, height = 6.5)
} else {
  message("Skipping tract map: the 'sf' package is not installed in this environment.")
}
```
