---
title: "Seasonality of Suspicious Stops"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Load data

```{r load}
library(dplyr)
library(ggplot2)
source("R/clean_stop_data.R")
source("R/classify_stops.R")

data <- clean_stop_data()
```

## Identify suspicious stops

A stop is considered "suspicious" when `STOP_REASON_NONTICKET` includes
"Suspicion of criminal activity (self-initiated)" among its (possibly
multi-valued, comma-separated) reasons.

```{r flag-suspicious}
suspicious <- flag_suspicious_stops(data)
```

## Classify each stop by outcome

Each suspicious stop is assigned to exactly one outcome, in priority order:
an **Arrest** if `ARREST_CHARGES` is recorded, else a **Ticket** if
`TICKET_COUNT` is a positive number, else **No Charge**.

```{r classify-outcome}
suspicious <- classify_outcome(suspicious)
```

## Count suspicious stops per month, by outcome

```{r count-per-month}
suspicious_per_month <- suspicious %>%
  group_by(MONTH, OUTCOME) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(MONTH, OUTCOME)

suspicious_per_month %>%
  tidyr::pivot_wider(names_from = OUTCOME, values_from = count, values_fill = 0)
```

## Plot seasonality by outcome

```{r plot-seasonality}
outcome_colors <- c(
  "Arrest"    = "#2a78d6",
  "Ticket"    = "#eb6834",
  "No Charge" = "#1baf7a"
)

seasonality_plot <- suspicious_per_month %>%
  mutate(MONTH = as.Date(paste0(MONTH, "-01"))) %>%
  ggplot(aes(x = MONTH, y = count, color = OUTCOME)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  scale_color_manual(values = outcome_colors, name = "Outcome") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    title = "Suspicious Stops per Month, by Outcome",
    subtitle = "STOP_REASON_NONTICKET includes 'Suspicion of criminal activity (self-initiated)'",
    x = "Month",
    y = "Number of Stops"
  )

seasonality_plot
```

```{r save-plot}
ggsave("plots/suspicious_stops_seasonality.png", plot = seasonality_plot, width = 10, height = 5)
```

## Breakdown of arrests by charge category

`ARREST_CHARGES_CATEGORY` can list multiple categories per arrest
(comma-separated, e.g. `"Narcotics, Weapon Violations"` — 44% of arrests,
1,470 of 3,351, list more than one). List order in this field is
alphabetical, not severity or chronology, so taking a single "primary"
category would arbitrarily favor categories earlier in the alphabet.
Instead, each arrest is exploded into one row per category it lists, so an
arrest charged with both Narcotics and Weapon Violations adds one count to
*each*. Monthly category totals can therefore exceed the number of arrests
that month. The top 7 categories by total count are kept individually (7 is
the most the validated color palette supports alongside an "Other" series,
8 total); the rest are folded into "Other" to keep the chart readable. With
only the top 5 broken out, categories like Traffic Violations and Other
Crimes (the 6th and 7th largest) were being absorbed into "Other" even
though they're individually comparable in size to the categories shown, so
"Other" swelled to dominate the chart; breaking those two out as their own
series flattens "Other" back down to a genuine long tail.

```{r classify-charge-category}
# read.csv() (unlike readr) doesn't trim whitespace, so a raw split on "," would
# leave a leading space on every list's first item (" Narcotics" vs "Narcotics"),
# silently splitting each category's count across two distinct strings.
arrests <- suspicious %>%
  filter(OUTCOME == "Arrest") %>%
  mutate(CHARGE_CATEGORY_LIST = lapply(strsplit(ARREST_CHARGES_CATEGORY, ","), trimws)) %>%
  tidyr::unnest(CHARGE_CATEGORY_LIST)

top_categories <- arrests %>%
  count(CHARGE_CATEGORY_LIST, sort = TRUE) %>%
  slice_head(n = 7) %>%
  pull(CHARGE_CATEGORY_LIST)

arrests <- arrests %>%
  mutate(
    CHARGE_CATEGORY = ifelse(CHARGE_CATEGORY_LIST %in% top_categories,
      CHARGE_CATEGORY_LIST, "Other"
    ),
    CHARGE_CATEGORY = factor(CHARGE_CATEGORY, levels = c(top_categories, "Other"))
  )
```

```{r count-charge-category-per-month}
charge_category_per_month <- arrests %>%
  group_by(MONTH, CHARGE_CATEGORY) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(MONTH, CHARGE_CATEGORY)

charge_category_per_month %>%
  tidyr::pivot_wider(names_from = CHARGE_CATEGORY, values_from = count, values_fill = 0)
```

```{r plot-charge-category}
charge_category_colors <- setNames(
  c("#2a78d6", "#eb6834", "#1baf7a", "#eda100", "#e87ba4", "#008300", "#4a3aa7", "#e34948"),
  c(top_categories, "Other")
)

charge_category_plot <- charge_category_per_month %>%
  mutate(MONTH = as.Date(paste0(MONTH, "-01"))) %>%
  ggplot(aes(x = MONTH, y = count, color = CHARGE_CATEGORY)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  scale_color_manual(values = charge_category_colors, name = "Charge Category") +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.subtitle = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 3)) +
  labs(
    title = "Arrests from Suspicious Stops per Month, by Charge Category",
    subtitle = stringr::str_wrap(
      "Top 7 categories from ARREST_CHARGES_CATEGORY; an arrest with multiple charges counts toward each of its categories, so monthly totals can exceed the arrest count; remainder folded into 'Other'",
      width = 95
    ),
    x = "Month",
    y = "Number of Arrests"
  )

charge_category_plot
```

```{r save-charge-category-plot}
ggsave("plots/suspicious_stops_arrest_charge_category.png", plot = charge_category_plot, width = 10, height = 5.5)
```
